---
title: "NH3 Cattle"
author: "Lorenzo Benedetti"
date: "2026-03-17"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)

rm(list = ls())

setwd("/Users/lorenzobenedetti/Documents/GitHub/Benedetti-2026-N2O-AgroscenaNext/Rstudio/Dynamic chambers")

library(readxl)
library(dplyr)
library(ggplot2)
library(car)
library(emmeans)
library(performance)
library(knitr)
library(kableExtra)
```

# Data import and preparation

The Excel file must contain the variables `TREATMENT`, `BLOCK`, and
`N-NH3 perc TAN`. Each row must represent one independent experimental unit.

```{r import-data}
data_file <- "Cattle Final NH3.xlsx"

if (!file.exists(data_file)) {
  stop("The file 'Cattle Final NH3.xlsx' was not found in the working directory.")
}

data <- read_excel(data_file)

required_columns <- c("TREATMENT", "BLOCK", "N-NH3 perc TAN")
missing_columns <- setdiff(required_columns, names(data))

if (length(missing_columns) > 0) {
  stop(
    "The following columns are missing from the Excel file: ",
    paste(missing_columns, collapse = ", ")
  )
}

data <- data %>%
  rename(NH3_TN = `N-NH3 perc TAN`) %>%
  mutate(
    TREATMENT = factor(TREATMENT),
    BLOCK = factor(BLOCK)
  )

if (anyNA(data[c("NH3_TN", "TREATMENT", "BLOCK")])) {
  warning(
    "Missing values are present in NH3_TN, TREATMENT, or BLOCK. ",
    "The model will use complete observations only."
  )
}
```

# Experimental-design check

The following table shows the number of observations for each block-by-treatment
combination. In a randomized complete block design, each cell should contain
one observation.

```{r design-check}
design_table <- with(data, table(BLOCK, TREATMENT, useNA = "ifany"))

kable(
  as.data.frame.matrix(design_table),
  caption = "Number of observations per block and treatment"
) %>%
  kable_styling(full_width = FALSE, position = "center")

if (any(design_table != 1)) {
  warning(
    "The design does not contain exactly one observation for each ",
    "BLOCK x TREATMENT combination. Because Type I sums of squares are ",
    "sequential, verify the data and model before interpreting the results."
  )
}
```

# Descriptive statistics

The following values are unadjusted descriptive statistics. Block-adjusted
estimated marginal means are reported below.

```{r descriptive-statistics}
desc <- data %>%
  group_by(TREATMENT) %>%
  summarise(
    mean = mean(NH3_TN, na.rm = TRUE),
    sd = sd(NH3_TN, na.rm = TRUE),
    n = sum(!is.na(NH3_TN)),
    se = sd / sqrt(n),
    .groups = "drop"
  )

kable(
  desc,
  digits = 3,
  caption = "Descriptive statistics by treatment"
) %>%
  kable_styling(full_width = FALSE, position = "center")
```

# Statistical model

Cumulative ammonia emissions are analysed according to a randomized complete
block design using the model

$$
Y_{ij} = \mu + \beta_j + \tau_i + \varepsilon_{ij},
$$

where $\beta_j$ is the fixed effect of block, $\tau_i$ is the fixed effect of
treatment, and $\varepsilon_{ij}$ is the residual error. Residuals are assumed
to be independent and normally distributed with a mean of zero and constant
variance. A block-by-treatment interaction is not included because only one
observation is available for each block-by-treatment combination.

```{r linear-model}
# BLOCK is deliberately entered before TREATMENT.
model <- lm(NH3_TN ~ BLOCK + TREATMENT, data = data)

coefficient_table <- as.data.frame(summary(model)$coefficients)

kable(
  coefficient_table,
  row.names = TRUE,
  digits = 4,
  caption = "Coefficients of the linear model including block and treatment"
) %>%
  kable_styling(full_width = FALSE, position = "center")
```

## Sequential Type I analysis of variance

Sequential Type I sums of squares are used. Because `BLOCK` is entered before
`TREATMENT`, the treatment effect is tested after accounting for block. In a
complete balanced block design, the treatment test is identical to the
corresponding Type II test.

```{r type-I-anova}
anova_type1 <- anova(model)

kable(
  as.data.frame(anova_type1),
  row.names = TRUE,
  digits = 4,
  caption = "Sequential Type I ANOVA: block entered before treatment"
) %>%
  kable_styling(full_width = FALSE, position = "center")

model_information <- data.frame(
  observations = nobs(model),
  residual_df = df.residual(model),
  residual_sd = sigma(model),
  R_squared = summary(model)$r.squared,
  adjusted_R_squared = summary(model)$adj.r.squared
)

kable(
  model_information,
  digits = 4,
  caption = "Model information"
) %>%
  kable_styling(full_width = FALSE, position = "center")
```

# Estimated marginal means

Estimated marginal means are adjusted for the block effect and are reported
with 95% confidence intervals.

```{r estimated-marginal-means}
emm <- emmeans(model, ~ TREATMENT)

emm_results <- summary(
  emm,
  infer = c(TRUE, FALSE),
  level = 0.95
)

kable(
  as.data.frame(emm_results),
  digits = 3,
  caption = "Block-adjusted estimated marginal means with 95% confidence intervals"
) %>%
  kable_styling(full_width = FALSE, position = "center")
```

# Tukey-adjusted treatment contrasts

All pairwise comparisons are performed on the block-adjusted estimated marginal
means. Tukey's method is used to adjust both the P values and simultaneous 95%
confidence intervals. No Fisher's least significant difference test is used.
The estimated mean differences and their confidence intervals are the primary
effect sizes and measures of uncertainty.

```{r tukey-contrasts}
tukey_contrasts <- pairs(emm, adjust = "tukey")

contrast_results <- summary(
  tukey_contrasts,
  infer = c(TRUE, TRUE),
  level = 0.95,
  adjust = "tukey"
)

kable(
  as.data.frame(contrast_results),
  digits = 3,
  caption = paste(
    "Estimated treatment differences with simultaneous 95% confidence intervals",
    "and Tukey-adjusted P values"
  )
) %>%
  kable_styling(full_width = FALSE, position = "center")
```

# Model diagnostics

Diagnostic tests have limited power with a small number of experimental units;
therefore, their results should be interpreted together with the residual and
influence plots.

```{r diagnostic-tests}
shapiro_res <- shapiro.test(residuals(model))

shapiro_table <- data.frame(
  test = "Shapiro-Wilk test on model residuals",
  statistic = unname(shapiro_res$statistic),
  p_value = shapiro_res$p.value
)

diagnostic_data <- data.frame(
  residual = residuals(model),
  treatment = model.frame(model)$TREATMENT
)

levene_res <- car::leveneTest(
  residual ~ treatment,
  data = diagnostic_data,
  center = median
)

kable(
  shapiro_table,
  digits = 4,
  caption = "Shapiro-Wilk test of residual normality"
) %>%
  kable_styling(full_width = FALSE, position = "center")

kable(
  as.data.frame(levene_res),
  row.names = TRUE,
  digits = 4,
  caption = "Levene's test applied to the model residuals"
) %>%
  kable_styling(full_width = FALSE, position = "center")
```

```{r base-diagnostic-plots, fig.cap="Diagnostic plots for the linear model"}
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))
```

```{r performance-diagnostic-plots, fig.cap="Graphical assessment of model assumptions"}
check_model(model)
```

# Data representation by block

Lines connect observations belonging to the same block, showing the matched
structure of the experimental design.

```{r block-plot, fig.cap="Observed treatment values distinguished by block"}
ggplot(
  data,
  aes(
    x = TREATMENT,
    y = NH3_TN,
    group = BLOCK,
    colour = BLOCK
  )
) +
  geom_line(linewidth = 0.7, alpha = 0.7) +
  geom_point(size = 3) +
  theme_bw() +
  labs(
    x = "Treatment",
    y = "N-NH3 (% of TAN)",
    colour = "Block",
    title = "N-NH3 percentage of TAN by treatment and block"
  )
```

# Estimated marginal means and confidence intervals

```{r emm-plot, fig.cap="Block-adjusted estimated marginal means and 95% confidence intervals"}
emm_plot_data <- as.data.frame(emm_results)

ggplot(
  emm_plot_data,
  aes(x = TREATMENT, y = emmean)
) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.15
  ) +
  theme_bw() +
  labs(
    x = "Treatment",
    y = "Estimated marginal mean of N-NH3 (% of TAN)",
    title = "Block-adjusted treatment means with 95% confidence intervals"
  )
```

# R session information

```{r session-information}
sessionInfo()
```
